Statistical quantification of confounding bias in predictive modelling

### 2021-10-31 manuscript.tex

While various data cleaning methods might help in mitigating confounding bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, johnson2007adjusting}, it is often unclear which variables should be considered as confounders and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.

@@

Recently, two different approaches were proposed for quantifying confounding bias \citep{chaibub2019permutation, ferrari2020measuring}. However, these methods either fail to control type I error (as known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in ref.\cite{chaibub2019permutation}), or do not provide p-values at all\cite{ferrari2020measuring}. 
Moreover, without some modifications, they are only applicable for categorical variables and involve re-fitting the model, which might not be feasible for models with high computational cost (e.g. when trained with nested cross-validation).

@@

As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models (mnlogit)\cite{bennett1966multiple, jones1975proability}. The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except by extreme non-normality). This result confirms that the tests can be deployed in a wide variety of predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.

@@

 Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$). In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.

@@

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and might largely accelerate the development of clinically useful machine learning biomarkers.

@@

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Assessing the simple bivariate (unconditioned) dependence ($H0: \yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounding bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, in this case the proposed variety of the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

>>>>

While various data cleaning methods may help in mitigating confounding bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, johnson2007adjusting}, it is often unclear which variables should be considered as confounders and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.

@@

Recently, two different approaches were proposed for quantifying confounding bias \citep{chaibub2019permutation, ferrari2020measuring}. However, these methods either fail to control type I error (as known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in ref.\cite{chaibub2019permutation}), or do not provide p-values at all\cite{ferrari2020measuring}. 
Moreover, without some modifications, they are only applicable for categorical variables and involve re-fitting the model, which may not be feasible for models with high computational cost (e.g. when trained with nested cross-validation).

@@

As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable may already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models (mnlogit)\cite{bennett1966multiple, jones1975proability}. The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

@@

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except by extreme non-normality). This result confirms that the tests can be deployed in a wide variety of predictive modelling scenarios. While different biomedical applications may consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.

@@

 Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$). In other words, the test accepted the null hypothesis of full model bias, suggesting that this model may be, in fact, almost exclusively driven by motion artifacts.
This example highlights that the full confounder test may become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.

@@

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and may largely accelerate the development of clinically useful machine learning biomarkers.

@@

Obviously, a strong association between $\yhat$ and $\c$ may indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Assessing the simple bivariate (unconditioned) dependence ($H0: \yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounding bias of predictive models. Option 1, i.e. full confounder testing, may be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, in this case the proposed variety of the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
